Wine Quality Analysis by Esha Chaudhary
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(reshape2)
library(dplyr)
library(gridExtra)
library(corrplot)
```

```{r echo=FALSE, Load_the_Data}
# Load the Dataset
wine <- read.csv('wineQualityReds.csv')
```

This dataset is related to red variants of the Portuguese "Vinho Verde" wine. The objective is to explore the dataset and see what properties influenced the quality of red wine. 

#Summary Statistics
```{r echo=FALSE}
str(wine)
```
```{r echo=FALSE, SUMMARY}
summary(wine)
```

There are 1599 observations and 13 variables. Variable X seems to be a unique identifier and won't help much in exploring the data. 
Quality which is the output variable has scores between 0 and 10. Mean and median of variables are also quite close to each other.

# Univariate Plots Section

I will start with plotting histogram for each variable to get an overview of the dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
#function to plot histograms for different variables in dataset
hist <- function(varname, df, bw, fillclr, xtitle){
  ggplot(df, aes_string(x = varname)) +
    geom_histogram(binwidth = bw, fill = fillclr) +
    xlab(xtitle)
}
```

1. Quality

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot  for quality
ggplot(data = wine, aes(x = quality)) +
  geom_histogram(fill = 'orange') + 
  xlab('Quality')
```
```{r echo=FALSE}
#stats for quality
summary(wine$quality)
```

```{r echo=FALSE}
#create new variable rating describing wine quality as bad, good, average.
wine$rating <- ifelse(wine$quality <= 3, 'bad', 
                    ifelse(wine$quality <= 6  , 'average', 
                           ifelse(wine$quality >= 7, 'good', 'none')))

wine$rating <- ordered(wine$rating, levels = c('bad', 'average', 'good'))

qplot(wine$rating, xlab="quality", fill = I('orange'))
```
```{r echo=FALSE}
by(wine$quality, wine$rating, summary)
```

Most of the wines in the dataset have average rating of 5 or 6.

2. Fixed Acidity

```{r echo=FALSE, message=FALSE, warning=FALSE}
#plot  for fixed acidity
hist('fixed.acidity',wine,0.2,'orange','Fixed acidity(g/l)') + 
  scale_x_continuous(breaks = seq(4,16,1))
```
```{r echo=FALSE}
summary(wine$fixed.acidity)
```

Fixed acidity have peaks at around 7 and 8. There are few outliers around 15 and 16.

3. Volatile Acidity

```{r echo=FALSE, message=FALSE, warning=FALSE}
#histogram for volatile acidity
hist('volatile.acidity',wine,0.02,'orange','Volatile acidity(g/l)') + 
  scale_x_continuous(breaks = seq(0,2,0.1))
```
```{r echo=FALSE}
summary(wine$volatile.acidity)
```

This plot has more than one peaks between 0.3 and 0.6.

4. Citric Acid

```{r echo=FALSE, message=FALSE, warning=FALSE}
#histogram for citric acid
hist('citric.acid',wine,0.01,'orange','Citric acid(g/l)') + 
  scale_x_continuous(breaks = seq(0,1,0.10))
```
```{r echo=FALSE}
summary(wine$citric.acid)
```

In this plot there are two peaks one at 0 and another at around 0.49. There are 132 wines in the dataset with no citric acid.

5. Residual Sugar

```{r echo=FALSE, message=FALSE, warning=FALSE}
#histograms for residual sugar
p1 <- hist('residual.sugar',wine,0.1,'orange','Sugar(g/l)') + 
  scale_x_continuous(breaks = seq(0,16,1))

p2 <- ggplot(data = wine, aes(x = residual.sugar)) +
  geom_histogram(fill = 'orange') +
  scale_x_log10() +
  xlab('log10(residual.sugar)')

grid.arrange(p1,p2)
```
```{r echo=FALSE}
summary(wine$residual.sugar)
```
This plot has a very long tail.

6. Chlorides

```{r echo=FALSE, message=FALSE, warning=FALSE}
#histogram for chlorides
p1 <- hist('chlorides',wine,0.01,'orange','chlorides(g/l)') + 
  scale_x_continuous(breaks = seq(0,1,0.1))

p2 <- ggplot(data = wine, aes(x = chlorides)) +
  geom_histogram(fill = 'orange') +
  scale_x_log10() +
  xlab('log10(chlorides)')

grid.arrange(p1,p2)
```
```{r echo=FALSE}
summary(wine$chlorides)
```
Chlorides have similar plot distribution as residual sugar.

7. Free Sulfur Dioxide

```{r echo=FALSE, message=FALSE, warning=FALSE}
#histogram for free sulfur dioxide
hist('free.sulfur.dioxide',wine,1,'orange','Free sulfur dioxide(mg/l)') + 
  scale_x_continuous(breaks = seq(1,100,5))
```
```{r echo=FALSE}
summary(wine$free.sulfur.dioxide)
```

Free sulfur dioxide has a peak value around 6 and is right skewed. 

8. Total Sulfur Dioxide

```{r echo=FALSE, message=FALSE, warning=FALSE}
#histogram total sulfur dioxide
hist('total.sulfur.dioxide',wine,1,'orange','total sulfur dioxide(mg/l)') + 
  scale_x_continuous(breaks = seq(1,300,20))

```

```{r echo=FALSE}
summary(wine$total.sulfur.dioxide)
```
Total sulfur dioxide has similar distribution as free sulfur. This makes sense as both are related.

9. Alcohol

```{r echo=FALSE, message=FALSE, warning=FALSE}
#histograms for alcohol
hist('alcohol',wine,0.1,'orange','alcohol(% by volume)') + 
  scale_x_continuous(breaks = seq(1,15,0.5))
```

```{r echo=FALSE}
summary(wine$alcohol)
```
Alcohol shows right skewed distribution.

10. Sulphates

```{r echo=FALSE, message=FALSE, warning=FALSE}
#histograms for sulphates
p1 <- hist('sulphates',wine,0.01,'orange','sulphates(g/l)') + 
  scale_x_continuous(breaks = seq(0,2,0.1))

p2 <- ggplot(data = wine, aes(x = sulphates)) +
  geom_histogram(fill = 'orange') +
  scale_x_log10()

grid.arrange(p1,p2)
```
```{r echo=FALSE}
summary(wine$sulphates)
```

Sulphates have normalish kind of distribution on logarithmic scale

11. pH

```{r echo=FALSE, message=FALSE, warning=FALSE}
#histogram for pH
hist('pH',wine,0.01,'orange','pH') + 
  scale_x_continuous(breaks = seq(0,5,0.1))
```

```{r echo=FALSE}
summary(wine$pH)
```
12. Density

```{r echo=FALSE, message=FALSE, warning=FALSE}
#histogram for density
hist('density',wine,0.0001,'orange','density(g/cm^3)')
```

```{r echo=FALSE}
summary(wine$density)
```
pH and density have normal distribution.

# Univariate Analysis

### What is the structure of your dataset?

Red wine dataset has 1599 observations and 13 variables. Quality varaible rates wine quality on scale from 0 to 10. Other attributes(fixed.acidity, volatile.acidity, citric.acid, residual.sugar, chlorides, free.sulfur.dioxide, total.sulfur.dioxide, density, pH, sulphates, alcohol) are all numerical variable. Attribute X is just a identifier for the record. 

### What is/are the main feature(s) of interest in your dataset?

Main feature in the dataset is the Quality. It's rating determines which wine is bad and which is good. 

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

Some features that are worth looking at are citric acid and volatile acid. One adds freshness while other gives unpleasant taste in wines. Alcohol and sulfur dioxide might affect quality too. We will analyze these variables in bivariate plots section.

### Did you create any new variables from existing variables in the dataset?

I created a rating variable which categorize wine quality as bad , average and good. 

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

Sugar, Chlorides and sulphates were visualized on logarithmic scale to get a better look at distribution of data. I also found 132 records with 0 value of citric acid.

# Bivariate Plots Section

I will start by plotting correlation matrix which will help in identifying variables that are related to each other.

```{r echo=FALSE, Bivariate_Plots}
corrplot(cor(wine[,2:13]), method="number",type = 'lower')
```

Corelation plot shows positive relationship of wine quality with alcohol and to some extent with sulphates and citric acid. Quality also possess negative relationship with volatile acidity. Free sulfur and total sulfur dioxide are highly related.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Function to create boxplots
plotbox <- function(x.var,y.var, df, xtitle,ytitle,fill.var){
  ggplot(df, aes_string(x = x.var, y = y.var, fill = fill.var)) +
    geom_boxplot() +
    scale_x_continuous(breaks = seq(3,8,1)) +
    xlab(xtitle) +
    ylab(ytitle)
}
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Boxplot of wine quality with alcohol
plotbox('quality','alcohol',wine,'Wine Quality','Alcohol(% by volume)','factor(quality)')
```

```{r echo=FALSE}
by(wine$alcohol,wine$quality,summary)
```
Good wines have higher alcohol percentage. Some wines have high alcohol content but their rating is average. There might be some other factors that would have degrated the quality of these wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#boxplot of quality vs citric acid
plotbox('quality','citric.acid',wine,'Wine Quality','Citric Acid(g/l)','factor(quality)')
```

```{r echo=FALSE}
by(wine$citric.acid,wine$quality,summary)
```
The boxplot shows median for citric acid increases with quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Boxplot quality vs sulpahtes
plotbox('quality','sulphates',wine,'Wine Quality','Sulphates(g/l)','factor(quality)')
```
```{r echo=FALSE}
by(wine$sulphates,wine$quality,summary)
```
Average wines have many outliers. This might have affected the correlation between quality and sulphates.

```{r echo=FALSE, message=FALSE, warning=FALSE}
plotbox('quality','volatile.acidity',wine,'Wine Quality','Volatile Acid(g/l)','factor(quality)')
```
```{r echo=FALSE}
by(wine$volatile.acidity,wine$quality,summary)
```
High quality wines have less volatile acids. This is expected as high levels of volatile acids can lead to an unpleasant, vinegar taste in wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}
plotbox('quality','chlorides',wine,'Wine Quality','Chlorides(g/l)','factor(quality)')
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
plotbox('quality','residual.sugar',wine,'Wine Quality','residual.sugar(g/l)','factor(quality)')
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
plotbox('quality','free.sulfur.dioxide',wine,'Wine Quality','free.sulfur.dioxide(mg/l)','factor(quality)')
```


Chlorides, residual sugar, free sulphur dioxide had very long tails. On plotting them using boxplot against quality we see that there are many outliers. Also there is no consistent relationship between these variables and wine quality.


Besides quality, there are some other variables that share quite strong relationship.
    Fixed acidity vs citric acid (0.67)
    Volatile acidity vs citric acid (-0.55)
    Fixed acidity vs density (0.67)
    Fixed acidity vs pH (-0.68)
    Citric acid vs pH (-0.54)
    Free sulfur dioxide vs total sulfur dioxide (0.67)
    alcohol and density(-0.5)
    density and pH(-0.34)


```{r echo=FALSE, message=FALSE, warning=FALSE}
p1 <- ggplot(aes(x = citric.acid, y = fixed.acidity), data = wine) +
  geom_jitter(alpha = 1/5) + 
  geom_smooth(method ='lm', color = 'red')

p2 <- ggplot(aes(x = citric.acid, y = volatile.acidity), data = wine) +
  geom_jitter(alpha = 1/5) + 
  geom_smooth(method ='lm', color = 'red')

p3 <- ggplot(aes(x = fixed.acidity, y = density), data = wine) +
  geom_jitter(alpha = 1/5) + 
  geom_smooth(method ='lm', color = 'red')

p4 <- ggplot(aes(x = fixed.acidity, y = pH), data = wine) +
  geom_jitter(alpha = 1/5) + 
  geom_smooth(method ='lm', color = 'red')

p5 <- ggplot(aes(x = citric.acid, y = pH), data = wine) +
  geom_jitter(alpha = 1/5) + 
  geom_smooth(method ='lm', color = 'red')

p6 <- ggplot(aes(x = total.sulfur.dioxide, y = free.sulfur.dioxide), data = wine) +
  geom_jitter(alpha = 1/5) + 
  geom_smooth(method ='lm', color = 'red')

p7 <- ggplot(aes(x = alcohol, y = density), data = wine) +
  geom_jitter(alpha = 1/5) + 
  geom_smooth(method ='lm', color = 'red')

p8 <- ggplot(aes(x = density, y = pH), data = wine) +
  geom_jitter(alpha = 1/5) + 
  geom_smooth(method ='lm', color = 'red')
grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8)

```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?
Quality strongly correlates with alcohol (0.48), volatile acidity (-0.39), sulphates (0.25) and citric acid (0.23). 

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?
Citric, volatile and fixed are closely related. This is also proved when we plot each of these against pH. pH describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale.

### What was the strongest relationship you found?
I found Alcohol as an important factor in predicting the quality. There are other variables also that shares strong relationship but alcohol directly affects the quality and hence is important in predicting the rating of wines.

# Multivariate Plots Section
In this part my focus will be only on variables that affects the quality of wines when combined.

```{r echo=FALSE, Multivariate_Plots}
#scatterplot b/w citric acid and volatile acidity vs quality
ggplot(data = wine,
aes(x = citric.acid, y = volatile.acidity,
    color = factor(quality))) +
  scale_color_brewer(type='seq',palette = 'OrRd') +
  geom_point() +
  facet_wrap(~rating) +
  theme_bw() +
  xlab('Citric acid(g/l)') +
  ylab('Volatile acid(g/l)')
```

```{r echo=FALSE}
#scatterplot b/walcohol and volatile acidity vs quality
ggplot(data = wine,
aes(x = alcohol, y = volatile.acidity,
    color = factor(quality))) +
  scale_color_brewer(type='seq',palette = "OrRd") +
  geom_point() +
  facet_wrap(~rating) +
  theme_bw() +
  xlab('Alcohol(% by volume)') +
  ylab('Volatile acid(g/l)')
```
```{r echo=FALSE}
#scatterplot b/walcohol and citric acidity vs quality
ggplot(data = wine,
aes(x = alcohol, y = citric.acid,
    color = factor(quality))) +
  scale_color_brewer(type='seq',palette = "OrRd") +
  geom_point() +
  facet_wrap(~rating) +
  theme_bw() +
  xlab('Alcohol(% by vol)') +
  ylab('Citric acid(g/l)')
```


```{r echo=FALSE}
#scatterplot b/w alcohol and sulpahtes vs quality
ggplot(data = wine,
aes(x = alcohol, y = sulphates,
    color = factor(quality))) +
  scale_color_brewer(type='seq',palette = "OrRd") +
  geom_point() +
  facet_wrap(~rating) +
  theme_bw() +
  xlab('Alcohol(% by vol)') +
  ylab('Sulphates(g/l)')
```


# Multivariate Analysis

### Talk about some of the relationships you observed in this part.

Above plots clearly shows that good wines have high level of alcohol and low level of volatile acids. Sulphates in average and good wines is more as compared to low rating wines.

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
plotbox('quality','alcohol',wine,'Wine Quality','Alcohol(% by volume)','factor(quality)')
```

### Description One

During my analysis the most important factor that I found affecting the quality of wines is alcohol. Correlation value between alcohol and quality is 0.47 which is highest as compared to other variables. Above plot clearly shows the median value for good wines to be quite high as compared to average and bad quality wines.

### Plot Two
```{r echo=FALSE, Plot_Two}
plotbox('quality','volatile.acidity',wine,'Wine Quality','Volatile Acid(g/l)','factor(quality)')
```

### Description Two

Volatile acidity is the second important factor that affects the quality of wines. It has a correlation value of -0.39. Median of good wines is lower. High amount of volatile acids in wine leads to an unpleasant taste and thus it makes sense to have it in low amount for wine to be taste good.

### Plot Three
```{r echo=FALSE, Plot_Three}
#scatterplot b/walcohol and volatile acidity vs quality
ggplot(data = wine,
aes(x = alcohol, y = volatile.acidity,
    color = factor(quality))) +
  scale_color_brewer(type='seq',palette = "OrRd") +
  geom_point() +
  facet_wrap(~rating) +
  theme_bw()+
  xlab('Alcohol(% by volume)') +
  ylab('Volatile acid(g/l)')
```

### Description Three

Since I found alcohol and volatile acids to be two imortant factor , I wanted to see if their combined effect supports my analysis. From the above plot it is quite clear that Good wines tends to have high alcohol content and lower volatile acids.

# Reflection

I started my EDA by creating histograms for each attribute present in the dataset. I also created a new variable that categorized wines as bad, average and good. My main focus was on identifying the factors that affected the quality of wines. Through my analysis I found alcohol and volatile acids as two important factors in determining the quality of wines. However I believe this study could have been improved if we had some more data for bad and good wines. Most of the wines in the dataset belonged to average quality and there are very less samples of bad wines. 

